---
title: "Project2_DSC680_Mahajan_Nitin"
output: html_notebook
---

#Loading the libraries and dataset 
```{r}
library(dplyr)
library(Seurat)
library(patchwork)

# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "/Users/nitin_mahajan/Downloads/filtered_gene_bc_matrices 2/hg19")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```

# Quality Control and Pre-Processing 
```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```

# Visualization of QC Metrics 

# Filter cells that have unique feature counts over 2,500 or less than 200
# Filter cells that have >5% mitochondrial counts
```{r}
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

# Feature Feature relationship 
```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

# Normalizing the data (method “LogNormalize”) 

```{r}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

```{r}
pbmc <- NormalizeData(pbmc)
```

# Feature Selection (features that exhibit high cell-to-cell variation in the dataset)
# In this data set we slected the 2,000 features
# The resulted data will be used for the further data anaysis

```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 20 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

# Scaling the data
# Using the ScaleData() function,  a linear transformation (‘scaling’) was applied to the dataset. 
```{r}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

# Dimention Reduction - PCA 

```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```



```{r}
# Examine and visualize PCA results a few different ways
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```



```{r}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```


```{r}
DimPlot(pbmc, reduction = "pca")
```


# Heatmap - Primary sources of heterogeneity in a dataset
```{r}
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```



```{r}
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```



```{r}
# NOTE: This process can take a long time for big datasets, comment out for expediency. More
# approximate techniques such as those implemented in ElbowPlot() can be used to reduce
# computation time
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
```

# Visualization used for comparing the distribution of p-values for each PC with a uniform distribution (dashed line). 
```{r}
JackStrawPlot(pbmc, dims = 1:15)
```
# ALTERNATIVE METHOD -ELBOW PLOT - A ranking of principle components based on the percentage of variance explained by each one  In this example, we can observe an ‘elbow’ around PC9-10, suggesting that the majority of true signal is captured in the first 10 PCs.

```{r}
ElbowPlot(pbmc)
```


# Clustering the cells  
```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
```


# Look at cluster IDs of the first 10 cells
```{r}
# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 10)
```

# Non-linear dimensional reduction (UMAP/tSNE)
# The goal of these algorithms is to learn the underlying manifold of the data in order to place similar cells together in 
# low-dimensional space.
```{r}
pbmc <- RunUMAP(pbmc, dims = 1:10)
```

```{r}
DimPlot(pbmc, reduction = "umap")
```


# Finding differentially expressed features (cluster biomarkers) in cluster 2 
# Here I Show top 10 genes/feature differentially expressed with p values and log2 Fold Change 

```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2, min.pct = 0.25)
head(cluster2.markers, n = 10)
```

# Comparing one cluster to another one
# finding all markers distinguishing cluster 5 from clusters 0 and 3
```{r}
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)
head(cluster5.markers, n = 5)
```


# find markers for every cluster compared to all remaining cells, report only the positive ones
# Filtered by logfc more than 0.25
```{r}
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)
```


# Distribution of selected genes in the object
# One can select the genes of interest and see the expression level in the data set in all the indentfied clusters 
```{r}
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
```

# Instead of expression level one can also plot the counts
```{r}
# you can plot raw counts as well
VlnPlot(pbmc, features = c("NKG7", "PF4"), slot = "counts", log = TRUE)
```

# To see the distribution of sleected genes in the all cluster to see the distribution
```{r}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP",
    "CD8A"))
```


```{r}
pbmc.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

# Assigning cell type identity to clusters
As we have the prior knowlegde of the markers of the cell population, we can name and identify them based on the following makrers

Cluster ID	Markers	Cell Type
0	IL7R, CCR7	Naive CD4+ T
1	CD14, LYZ	CD14+ Mono
2	IL7R, S100A4	Memory CD4+
3	MS4A1	B
4	CD8A	CD8+ T
5	FCGR3A, MS4A7	FCGR3A+ Mono
6	GNLY, NKG7	NK
7	FCER1A, CST3	DC
8	PPBP	Platelet

```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono",
    "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

